Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-03-01 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd

In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded. For example, students may be clustered in classes in psychometrics research, or patients may be clustered in studies in individual patient data meta-analyses. Ignoring the clustered structure of such multilevel data can be harmful to the statistical inferences and bias results. Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present. The cluster to which a unit belongs may then influence the unit-level observations--and vice versa for each of the units that make up the cluster [@hox17]. These relations can and should be taken into account when developing analysis models for multilevel data for the simple reason that groups of observations share some common variance. The variability due to clustering is often measured by means of the intraclass coefficient [ICC, @fish25]. The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure. Multilevel models typically accommodate for this variability by including a separate group mean for each cluster. In addition to random intercepts, multilevel models can also include random effects and heterogeneous residual error variances across clusters [see e.g. @gelm06, @hox17 and @jong21]. There are many names for models that take clustering into account. Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.

@@

The process of analyzing multilevel data is further complicated when not all data entries are observed. Just as with single level data, missingness may occur at the unit level. But with multiple levels of data comes the potential for clustered missingness. Incomplete multilevel data can be categorized into two general patterns: systematic missingness and sporadic missingness [@resc13]. Systematic missingness implies that one or more variables are never observed in a certain cluster. With sporadic missingness there may be observed data for some but not all units in a cluster [@buur18; @jola18]. We have visualized this difference in Figure 1, which shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters, and $p=3$ columns. Column $X1$ is completely observed, column $X2$ is systematically missing, and column $X3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but need some kind of missing data method or discard the incomplete units altogether (i.e., list-wise deletion/complete case analysis). And, with the systematic missingness in $X2$, it is impossible to fit a multilevel model since we cannot estimate the intercept of cluster 2. We would have to exclude this cluster from our analyses entirely to obtain any results. Obviously, excluding observations is not a desirable workflow.

Ignoring the missingness in analyses can be extremely harmful to inferences. Complete case analysis can introduce bias in statistical inferences and lowers statistical power. Instead, the missingness should be accommodated \emph{before} or _within_ the analysis of scientific interest. Especially the former is very generic and popular. Imputing (i.e., filling in) the missing values separates the missing data problem from the scientific problem: missing data are replaced by plausible values whereafter the completed data is analyzed as if it were completely observed. The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis. \pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18]. In this paper, we will discuss how to use \pkg{mice} in the context of multilevel data.

@@

- `popmis` from the \pkg{mice} package (simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools);
- `hiv` from the \pkg{GJRM} package (simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions);
- `impact` from the \pkg{metamisc} package (empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies).

For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow. With the `popmis` data, we show how (not) to develop an imputation model. With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods. And with the `impact` data we provide an example of multivariate missingness in real-world data. Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.

@@

@@

```{r impact_pred}
quickpred(impact_NA)
```

>>>>

~~In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded.~~ Contemporary decision making is driven increasingly often and sometimes even entirely dictated by data. Although data relevant for decision makers were initially collected using small and well-designed studies, there has been a growing need to address more complex questions at a wider scale. This not only requires to collect larger amounts of data, but also from more diverse settings and populations. 
For example, [TD: introduce IPDMA here; eg. HIV studies Johanna is referring to]. Another example is the use of large databases with information from thousands or even millions of individuals from multiple cities, regions or even countries. Finally, a third situation arises when data are collected from multicenter studies [TD: maybe refer to the schools/classes example here]. For example, students may be clustered in classes in psychometrics research, or patients may be clustered in studies in individual patient data meta-analyses in medical research. A common characteristic in aforementioned examples is the presence of clustering. [TD: briefly explain what is clustering?]

Ignoring the clustered structure of such multilevel data can be harmful to the statistical inferences and introduce bias in estimators [@hox17]. 
~~Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present. The cluster to which a unit belongs may then influence the unit-level observations--and vice versa for each of the units that make up the cluster.~~ 
These relations can and should be taken into account when developing analysis models for multilevel data for the simple reason that groups of observations share some common variance.
~~The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.~~ 
Multilevel models typically accommodate for this variability by including a separate group mean for each cluster. In addition to random intercepts, multilevel models can also include random effects and heterogeneous residual error variances across clusters [see e.g. @gelm06, @hox17 and @jong21]. There are many names for models that take clustering into account. Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'. [TD: emphasize when/why we need to account for clustering in the analysis of clustered data. Why is the presence of clustering relevant when considering multiple imputation of missing data? e.g. distinction between systematically and sporadically missing data. But also: mechanisms of missing data (e.g. MCAR, MAR, MNAR) may differ between clusters. But also: relation between observed data may differ between clusters? When/ why should we avoid using traditional imputation methods? e.g. congeniality issues.]

@@

The process of analyzing multilevel data is further complicated when not all data entries are observed. Just as with single level data, missingness may occur at the unit level. But with multiple levels of data comes the potential for clustered missingness. Therefore, incomplete multilevel data can be categorized into two general patterns: systematic missingness and sporadic missingness [@resc13]. Systematic missingness implies that one or more variables are never observed in a certain cluster. With sporadic missingness there may be observed data for some but not all units in a cluster [@buur18; @jola18]. We have visualized this difference in Figure 1, which shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters, and $p=3$ columns. Column $X1$ is completely observed, column $X2$ is systematically missing in cluster 2, and column $X3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method or discard the incomplete units altogether (i.e., list-wise deletion/complete case analysis). Further, with the systematic missingness in $X2$, it is impossible to fit a multilevel model since we cannot estimate the intercept of cluster 2. We would have to exclude this cluster from our analyses entirely to obtain any results. Obviously, excluding observations is not a desirable workflow.

Ignoring the missingness in analyses can be extremely harmful to inferences. Complete case analysis can introduce bias in statistical inferences and lowers statistical power. Instead, the missingness should be accommodated \emph{before} or _within_ the analysis of scientific interest. Especially the former is very generic and popular. [VJ: add why multiple values are imputed.] Imputing (i.e., filling in) the missing values separates the missing data problem from the scientific problem: missing data are replaced by plausible values whereafter the completed data is analyzed as if it were completely observed. The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis. \pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18]. In this paper, we will discuss how to use \pkg{mice} in the context of multilevel data.

@@

- `popmis` from the \pkg{mice} package [simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools, @mice];
- `hiv` from the \pkg{GJRM} package [simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions, @GJRM];
- `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies, @metamisc].

For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow. With the `popmis` data, we show how (and how not) to develop an imputation model. With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods. With the `impact` data we provide an example of multivariate missingness in real-world data. Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.

@@

TODO: make a 'translation table' between model equation and lme4. Use betas instead of gamma's and mu's.

@@

# use glm with name as dummy
# only random effect on ct, not the other variables
# make forest plot of ct variable
# is there heterogen in association of ct with mort?
# is there any random effect on that variable at all?
# 
```

```{r impact_pred}
pred <- quickpred(impact_NA)
pred[pred == 1] <- 2
pred[, "name"] <- -2
pred["mort", ] <- 2
pred[, "mort"] <- 2
diag(pred) <- 0
pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
pred

meth <- make.method(impact_NA)
meth[meth != ""] <- "2l.pmm"
```

```{r imp_impact}
# imp_impact <- impact_NA %>%
#   mutate(name = as.integer(name), motor_score = as.numeric(motor_score)) %>%
#   mice::mice(., m = 2, maxit = 1, method = meth, predictorMatrix = pred)
# look at jomo for categorical variables?
# semi-cont with jomo is not ideal (schafer, '97) because you need 2-step approach
# pmm is better (more efficient) because it will still look for donors (maybe outside of cluster) based on predictive distance, even for very small clusters
# make assumptions of these methods explicit!
```
